\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}

### 2016-03-04 write_up/cricket.tex
close

None of this to say that there is actually an advantage in winning the toss, or that teams are able to successfully exploit any such advantage. For it may be impossible to predict well in advance the advantage of bowling or batting first. Or, it may be that teams squander the potential advantage by using bad heuristics to choose what they do.

To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches --- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions. We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.

\section*{Data}
Data are from 43,185 first-class men's cricket matches. It is a near census of the relevant population. We have data on all types of matches: domestic and international Twenty20s --- T20s and T20Is respectively, domestic and international one-dayers --- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches --- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches. In English county cricket, first-class matches last four days. Some first class matches last just a day. Others two days. Yet others three days. And till a particular point in history, a test match lasted as long as it was needed to finish a game. We elide over such differences.}

Of these matches, we do not have data on the toss for 2,807, or roughly 7\%, of the matches. The primary reason we don't have data on these matches is because the match was abandoned without play. We exclude data from these matches.

In limited overs cricket, a minimum number of overs must be bowled to establish a result. In a one-day match, for instance, each side must bat at least 20 overs for a result to be declared. In 706 matches, or roughly 1.7\% of the remaining matches, not enough overs were bowled to get a result. We again exclude these matches from our analysis. This leaves us with 39,672 matches. We analyze these data.

@@

We assume that the outcome of a toss is random. Conditional on the outcome of a toss being random, the effect of winning the toss can be attributed to the toss itself. Any decision made after the outcome of the toss is known, however, is `post-treatment.' In particular, the decision to bat or bowl first is made after accounting for the relative strengths and weaknesses vis-\`{a}-vis the competing team at that particular instance, and thus not independent of team attributes. Hence, conditioning on decision to bowl or bat first can induce bias in estimates of advantage of winning a toss. Thus, unlike \citet{dawson2009bat}, \citet{Saad2015}, we solely rely on the assumption that the outcome of a toss is random.

But before we exploit the design, we shed some light on the validity of the assumption. In particular, we assess whether the coin toss is somehow rigged, with the home side enjoying the rub of the green more often. For this analysis, we only get to exploit international matches as establishing which of the teams is the home team in local matches is somewhat arduous. Of the 5,591 international matches, the home team won the toss in 2,860 matches, or about 51.15\%. The probability that we will get as large a proportion by chance after tossing 5,591 coins is about 8.7\%, low but not eyebrow-raisingly so. Thus, for now, we choose to trust that the tosses are random.

A caveat about interpretation before we report the results. As we discuss in the introduction, we cannot estimate the actual advantage of winning a toss. We can only estimate the net observed advantage, which is the extent to which the teams capitalize on the potential advantage. With that, the results.

The team that wins the toss wins the match 2.9\% more often than the team that loses the toss. This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads.
% Also interpret it by comparing to probability of winning when playing with the worst placed team (not defined from last team)
This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team. Much of the variability follows expected patterns.

>>>>

None of this to say that there is actually an advantage in winning the toss, or that teams are able to successfully exploit any such advantage. For it may be impossible to predict well in advance the advantage of bowling or batting first. (If pre-match assessments of the pitch by media commentators are anything to go by, error in assessment of conditions is likely large.) Or, it may be that teams squander the potential advantage by using bad heuristics to choose what they do. For instance, teams may weigh outcomes from recent matches `too' heavily; e.g., a team that has, of late, won chasing may choose to chase even though pitch conditions favor batting first.

To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches--- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions. We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.

\section*{Data}
Data are from 44,224 first-class cricket matches. It is a near census of the relevant population.\footnote{Data excludes scheduled matches that were abandoned without the toss being conducted.} We have data on all types of matches: domestic and international Twenty20s --- T20s and T20Is respectively, domestic and international one-dayers --- List A and One-Day Internationals (ODI) respectively, and domestic and international multi-day matches --- First Class (FC) and Tests respectively.\footnote{There is a rich variety of first-class matches. In English county cricket, first-class matches last four days. Some first class matches last just a day. Others two days. Yet others three days. And till a particular point in history, a test match lasted as long as it was needed to finish a game. We elide over such differences.}

Of the 44,224 matches, 1,019 matches were abandoned without play. We exclude these matches. In another 1,376 matches, we do not have information on whether the team chose to bat or bowl after the toss. Informal inspection suggests that data are missing because no match was played. We excludes these matches as well.

In limited overs cricket, a minimum number of overs must be bowled to establish a result. In a one-day match, for instance, each side must bat at least 20 overs for a result to be declared. In 769 matches, or roughly 1.7\% of the total matches, not enough overs were bowled to get a result. We exclude these matches from our analysis. This leaves us with data from 41,060 matches. We analyze these data.

@@

We assume that the outcome of a toss is random. Conditional on the outcome of a toss being random, the effect of winning the toss can be attributed to the toss itself. Any decision made after the outcome of the toss is known, however, is `post-treatment.' In particular, the decision to bat or bowl first is made after accounting for the relative strengths and weaknesses vis-\`{a}-vis the competing team at that particular instance, and thus not independent of team attributes. Hence, conditioning on decision to bowl or bat first can bias estimates of advantage of winning a toss. Thus, unlike \citet{dawson2009bat}, \citet{Saad2015}, we solely rely on the assumption that the outcome of a toss is random.

But before we exploit the design, we shed some light on the validity of the assumption. In particular, we assess whether the coin toss is somehow rigged, with the home side enjoying the rub of the green more often. For this analysis, we only get to exploit international matches as establishing which of the teams is the home team in local matches is somewhat arduous. Of the 5,684 international matches for which we can match the country of the ground to the country of one of the teams, the home team won the toss in 2,892 matches, or about 50.87\%. The chance of getting as many wins by fluke after tossing 5,684 coins is about 18.91\%. The chance is low, but not eyebrow raisingly so.

But, rather than consider all home matches, we may instead want to only consider matches that are officiated only by home umpires --- the norm till 1992.\footnote{For more information on move to neutral umpires, see \href{http://www.espncricinfo.com/magazine/content/story/511175.html}{Neutral Umpires} by S. Rajesh on ESPNcricinfo.} In matches featuring umpires from only the home country, the team with `home umpire' advantage won toss nearly 51.9\% of the times. And the chance of getting a greater percentage of wins than 51.9\% in 2,965 is a shade less than 4\%. Thus, there is some reason to worry that the tosses are rigged. Any such rigging would bias estimates of advantage of winning the coin toss to the extent that it is correlated with ability. More plainly, if stronger teams win more tosses, estimates of the advantage of the coin toss would be inflated upwards. And vice versa, if otherwise. We, however, do not have good reasons to think that there is a correlation. So for now, we proceed as if the tosses are random.

Another caveat about interpretation before we present the results. As we discuss in the introduction, we cannot estimate the actual advantage of winning a toss. We can only estimate the net observed advantage, which is the extent to which the teams capitalize on the potential advantage. With that, the results.

The team that wins the toss wins the match 2.8\% more often than the team that loses the toss. This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads. This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team. Much of the variability follows expected patterns.
